Template for indicator documentation

### 2023-10-18 Nature_Index_product.Rmd
Suggestions for text; added anchors

```{r setup, include=FALSE}
library(NIflex)
library(magrittr)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

@@

## Introduction

@@

## About the underlying data

@@

### Representativity in time and space

@@

### Original units

@@

### Temporal coverage

@@

### Aditional comments about the dataset

@@

## Ecosystem characteristic

### Norwegain standard

@@

### SEEA EA (UN standard)

@@

## Collinearities with other indicators

@@

## Reference condition and values

### Reference condition

@@

### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values

As NI products are already scaled to range from 0 to 1 and reference conditions are set according to the same approach, the reference value is 1.
For wetlands ecosystems in particular, we have not yet discussed the definition of "good ecological condition", but it seems that in the Ecological Condition Assessments for forest and mountain ecosystems, 0.6 was used as "good ecological condition" consistent with the approach from "Vanndirektivet".
To what degree this is appropriate is difficult to judge.
It may well be best to use 0.6 due to its consistency with other frameworks (e.g. Vanndirektivet) as it would be challenging to otherwise determine what is "good enough" condition for a highly derived variables such as NI products.

## Uncertainties

Calculated NI products are actually composed of a pre-defined number of random samples from a probability distribution drawn during an internal bootstrap implemented in the R package NIcalc.
These draws can be summarised with metrics such as mean, median, confidence intervals.
etc.
The degree of uncertainty is determined by a) the uncertainty associated with each indicator's data as deposited in the NI database and b) the uncertainty resulting from imputation of missing values.

Besides these "quantitative" sources of uncertainty, there is also "qualitative" uncertainty linked to using NI products as indicators in Ecological Condition Assessments.
This is linked to methodological questions regarding how (and perhaps if) highly processed composite indices should be included as indicators to make other composite indices.
While some of these questions are of a more philosohpical nature, some more pragmatic ones include decisions about which indicators to include/exclude (especially considering indicators that may get used directly and via NI products and indicators that do not really quantify species abundance), how to weigh indicators, and whether or not to impute NA values as part of generating the NI product.

@@

## Suggestions for using NI products in Ecological Condition Assessments

@@

The general rule is that NI products used for ecosystem condition assessments should follow a definition of NI as an indicator of the abundance of species. This means that the indicators that are part of such NI products should be indicators representing abundance and/or population status of specific species and species groups. Indicators that describe properties linked to the condition and/or occurrence of habitat types, on the other hand, should be dropped. For wetlands, for example, this includes the NI indicators "Atlantisk høgmyr areal" and "Palsmyr areal". 
The second group of indicators that should not be included in NI products are indicators that are also used as additional independent indicators in the ecosystem assessment, or that use the same underlying data as other indicators in the ecosystem assessment. This includes, e.g., indicators such as breeding birds, wolves, small rodents etc. Including the same data/information through different channel is problematic for several reasons including artificially increased indicator weight, inflated precision of the resulting aggregate metrics, high correlations among indicators, and increased risk for mistakes due to paralell workflows.  
The last group of indicators from the NI database that needs discussing are indicators based on expert knowledge. It has been decided that ecosystem condition assessments using the IBECA method should only include indicators based on data and model predictions. To what degree NI products that contain a portion of indicators based on expert knowledge should be allowed will require discussion.
Considering all the groups of indicators that should not be included in NI products for ecosystem condition assessments opens up a discussion on whether, instead of using a composite index across a small and very unbalanced set of species (groups), a better approach might be to design NI products that represent specific taxonomic and/or functional species groups. Take, for example, the wetlands NI product presented below: if the bird indicators are removed as they are a separate index, and the four indicators representing expert knowledge on rare mosses and insects are discontinued, all that remains is eight species of vasular plants and wild reindeer. Instead of averaging across plants and reindeer here, it may be more sensible to make an NI product for vascular plants, and potentially include reindeer as a separate indicator.

Weighting of indicators during aggregation happens at three levels in the NI framework: area weighting, trophic weighting, and key indicator weighting. 
Area and trophic weighting aim to smooth over imbalances in spatial and taxonomic coverage of indicators while increased weights for a priori defined key indicators are supposed to increase the influence of indicators that are thought to be functionally important species and habitats. 
Area weighting is central to the method and should always be done when calculating NI products, while the need for trophic weighting depends on the indicator selection of the specific NI product: weighting by trophic group can be useful for indicator sets that are taxonomically diverse while it is largely moot for sets of indicators representing functionally and/or taxonomically closely related species. 
For the majority of NI products for inclusion in ecosystem condition assessments key indicator weighting will not be relevant because most NI key indicators fall into the categories of either habitat indicators and/or indicators that are used independently in the assessments and are therefore removed from NI products. If any key indicators remain in the relevant indicator set after that, there is a good chance it will be only one or two, and allocating literally half the weight in a composite index to a single indicator is not reasonable.

@@

## References

@@

## Analyses

The following analysis workflow is almost entirely arranged in functions.
The functions for the different steps are from the NIflex R package (<https://github.com/NINAnor/NIflex>).Many of the NIflex functions call on functions from NIcalc (<https://github.com/NINAnor/NIcalc>), the R package written for entering and extracting data from the NI database and calculating NI products from it.
Before proceeding any further, we therefore need to make sure that both of these packages and their dependencies are installed:

@@

NIflex was written to provide a flexible workflow for calculating a range of NI products including the original NI for all seven main ecosystems, a range of previously used thematic indices, fully custom indices and -- most important here -- NI products aimed for use in Ecological Condition Assessments for forest, mountain, and wetland ecosystems.

@@

### Data sets

#### Nature Index database

@@

#### Regions

@@

### Workflow options

NIflex offers a workflow for calculating NI products that can be customized in terms of a range of options.
So next, we set the global options for calculating our wetlands NI product, starting with the years of data we want to include.
Since we will not use NA imputation here, we drop the first year with available data for in the NI database (1990) as many indicators do not actually have values for this year.

@@

```{r}
Diagnostics <- TRUE # Yes
TestRun <- FALSE # Yes
```

@@

### Scaled indicator values

@@

We also need to set variables "theme" and "funArguments" to "none" and `NULL`, respectivel (these are used for controlling workflow elements linked to thematic indices):

@@

```{r, results = FALSE, eval=FALSE}
CustomNI <- calculateCustomNI(ecosystem = ecosystem_use,
                              indicators = indicators_use,
                              theme = theme,
                              dropInd = dropInd,
                              username = NIdb_username,
                              password = NIdb_password,
                              KeyIndicators = KeyIndicators,
                              KeyWeight = KeyWeight,
                              AreaWeights = AreaWeights,
                              TrophicWeights = TrophicWeights,
                              NAImputation = NAImputation,
                              years = years,
                              OutputType = OutputType,
                              funArguments = funArguments,
                              Diagnostics = Diagnostics,
                              TestRun = TestRun,
                              norwegianNames = norwegianNames,
                              saveSteps = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
CustomNI <- readRDS("data/cache/CustomNI.RDS")
```

@@

```{r, eval=FALSE}
CustomNI_map <- geomapNI(Index = CustomNI$CustomIndex, 
                         year = 2019, 
                         OutputType = OutputType, 
                         ecosystem = ecosystem_use,
                         shapeLibraryPath = shapeLibraryPath)
```

@@

```{r, eval=FALSE}
Index <- CustomNI$CustomIndex
plot(Index$wholeArea, main = "Custom index", cex = 1, lwd = 2, shade = TRUE)
summary(Index$wholeArea)
```

![](images/custPlot.png)

>>>>

```{r setup, include=FALSE}
library(NIflex)
library(magrittr)
library(knitr)
library(flextable)
knitr::opts_chunk$set(echo = TRUE)
```

@@

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT,
                       check.names = F)
knitr::kable(metaData)
```

@@

## Introduction {#intro-ni}

@@

## About the underlying data {#underlying-data-ni}

@@

### Representativity in time and space {#representativity-ni}

@@

### Original units {#units-ni}

@@

### Temporal coverage {#temporal-coverage-ni}

@@

### Aditional comments about the dataset {#comments-ni}

@@

## Ecosystem characteristic {#ecological-char-ni}

### Norwegain standard {#norwegian-standard-ni}

@@

### SEEA EA (UN standard) {#seea-ni}

@@

## Collinearities with other indicators {#collinear-ni}

@@

## Reference condition and values {#ref-con-ni}

### Reference condition {#ref-con2-ni}

@@

### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-values-ni}

As NI products are already scaled to range from 0 to 1 and reference conditions are set according to the same approach, the reference value is 1.
For wetlands ecosystems in particular, we have not yet discussed the definition of "good ecological condition", but in the Ecological Condition Assessments for forest and mountain ecosystems, the default 0.6 was used as "good ecological condition".
To what degree this is appropriate is difficult to judge.
It may well be best to use 0.6 due to its consistency previous work as it would be challenging to at least empirically determine what is "good enough" condition for a highly derived variables such as NI products.

## Uncertainties {#uncertainties-ni}

Calculated NI products are actually composed of a pre-defined number of random samples from a probability distribution drawn during an internal bootstrap implemented in the R package ``NIcalc``.
These draws can be summarised with metrics such as mean, median, confidence intervals.
etc.
The degree of uncertainty is determined by a) the uncertainty associated with each indicator's data as deposited in the NI database and b) the uncertainty resulting from imputation of missing values.

Besides these "quantitative" sources of uncertainty, there is also "qualitative" uncertainty linked to using NI products as indicators in Ecological Condition Assessments.
This is linked to methodological questions regarding how (and perhaps if) highly processed composite indices should be included as indicators to make other composite indices.
While some of these questions are of a more philosophical nature, some more pragmatic ones include decisions about which indicators to include/exclude (especially considering indicators that may get used directly and via NI products and indicators that do not really quantify species abundance), how to weigh indicators, and whether or not to impute NA values as part of generating the NI product.

@@

## Recommendations for using NI products in Ecological Condition Assessments {#recomendations-ni}

@@

The general rule is that NI products used for ecosystem condition assessments should follow a definition of NI as an indicator of the abundance of species. This means that the indicators that are part of such NI products should be indicators representing abundance and/or population status of specific species and species groups. Indicators that describe properties linked to the condition and/or occurrence of habitat types, on the other hand, should be dropped. For wetlands, for example, this includes the NI indicators "Atlantisk høgmyr areal" and "Palsmyr areal". 
The second group of indicators that should not be included in NI products are indicators that are also used as additional independent indicators in the ecosystem assessment, or that use the same underlying data as other indicators in the ecosystem assessment. This includes, e.g., indicators such as breeding birds, wolves, small rodents etc. Including the same data/information through different channels is problematic for several reasons including artificially increased indicator weight, inflated precision of the resulting aggregate metrics, high correlations among indicators, and increased risk for mistakes due to parallel workflows.  
The last group of indicators from the NI database that needs discussing are indicators based on expert knowledge. It has been decided that ecosystem condition assessments using the IBECA method should only include indicators based on data and model predictions. To what degree NI products that contain a portion of indicators based on expert knowledge should be allowed will require discussion. It is clear, however, that expert knowledge can also be considered data. There is also large variation in how expert elicitation is conducted which will influence the quality and transparency of the this kind of data.
Considering all the groups of indicators that should not be included in NI products for ecosystem condition assessments opens up a discussion on whether, instead of using a composite index across a small and very unbalanced set of species (groups), a better approach might be to design NI products that represent specific taxonomic and/or functional species groups. Take, for example, the wetlands NI product presented below: if the bird indicators are removed as they are a separate index, and the four indicators representing expert knowledge on rare mosses and insects are discontinued, all that remains are eight species of vascular plants and wild reindeer. Instead of averaging across plants and reindeer here, it may be more sensible to make an NI product for vascular plants, and potentially include reindeer as a separate indicator.

Weighting of indicators during aggregation happens at three levels in the NI framework: area weighting, trophic weighting, and key indicator weighting. 
Area and trophic weighting aim to smooth over imbalances in spatial and taxonomic coverage of indicators while increased weights for _a priori_ defined key indicators are supposed to increase the influence of indicators that are thought to be functionally important species or habitats. 
Area weighting is central to the method and should always be done when calculating NI products, while the need for trophic weighting depends on the indicator selection of the specific NI product: weighting by trophic group can be useful for indicator sets that are taxonomically diverse while it is largely moot for sets of indicators representing functionally and/or taxonomically closely related species. 
For the majority of NI products for inclusion in ecosystem condition assessments key indicator weighting will not be relevant because most NI key indicators fall into the categories of either habitat indicators and/or indicators that are used independently in the assessments and are therefore removed from NI products. If any key indicators remain in the relevant indicator set after that, there is a good chance it will be only one or two, and allocating literally half the weight in a composite index to a single indicator is not reasonable.

NI indicators are normalised (i.e. scaled relative to their reference values) at what is referred to as the BSU level, or basic spatial units. 
This is the spatial resolution of the raw observation data, and the reference values have the same spatial resolution as the observation data.
Quite often the BSU it is a municipal scale, but not always. 
The EAA (ecosystem assessment area), however, is typically much larger, with only five regions in Norway.
Therefore, scaled indicator valeus are aggregated to produce regional estimates, with the effect that the regional estimated now represents the (area weighted) condition of the BSUs, and not the staus of the population.
An alternative to this, for indicators where the raw data is in the form of counts or mass (densities are not suited), one might sum the variable and the reference values for each region and then scale the variable based on these sums (e.g. number of individuals).
This approach would produce an indicator which reflect the _population_ status rather than the mean status for any given area.

@@

## References {#refs.ni}

@@

## Analyses {#analyses-ni}

The following analysis workflow is almost entirely arranged in functions.
The functions for the different steps are from the ``NIflex`` R package (<https://github.com/NINAnor/NIflex>).Many of the NIflex functions call on functions from ``NIcalc`` (<https://github.com/NINAnor/NIcalc>), the R package written for entering and extracting data from the NI database and calculating NI products from it.
Before proceeding any further, we therefore need to make sure that both of these packages and their dependencies are installed:

@@

`NIflex`` was written to provide a flexible workflow for calculating a range of NI products including the original NI for all seven main ecosystems, a range of previously used thematic indices, fully custom indices and -- most important here -- NI products aimed for use in Ecological Condition Assessments for forest, mountain, and wetland ecosystems.

@@

### Data sets {#data-sets-ni}

#### Nature Index database {#ni-database-ni}

@@

#### Regions {#regaions.ni}

@@

### Workflow options {#workflow-options-ni}

`NIflex`` offers a workflow for calculating NI products that can be customized in terms of a range of options.
So next, we set the global options for calculating our wetlands NI product, starting with the years of data we want to include.
Since we will not use NA imputation here, we drop the first year with available data for in the NI database (1990) as many indicators do not actually have values for this year.

@@

```{r}
Diagnostics <- TRUE # Yes
TestRun <- FALSE # No
```

@@

### Scaled indicator values {#scaled-values-ni}

@@

We also need to set variables "theme" and "funArguments" to "none" and `NULL`, respectively (these are used for controlling workflow elements linked to thematic indices):

@@

```{r, results = FALSE, eval=FALSE}
CustomNI <- calculateCustomNI(ecosystem = ecosystem_use,
                              indicators = indicators_use,
                              theme = theme,
                              dropInd = dropInd,
                              username = NIdb_username,
                              password = NIdb_password,
                              KeyIndicators = KeyIndicators,
                              KeyWeight = KeyWeight,
                              AreaWeights = AreaWeights,
                              TrophicWeights = TrophicWeights,
                              NAImputation = NAImputation,
                              years = years,
                              OutputType = OutputType,
                              funArguments = funArguments,
                              Diagnostics = Diagnostics,
                              TestRun = TestRun,
                              norwegianNames = norwegianNames,
                              saveSteps = FALSE)
```
```{r, include=F}
dir <- substr(getwd(), 1,2)
pData <- ifelse(dir == "C:", 
               "P:/41201785_okologisk_tilstand_2022_2023/data/",
               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
saveRDS(CustomNI, paste0(pData, "cache/CustomNI.RDS"))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
CustomNI <- readRDS(paste0(pData, "cache/CustomNI.RDS"))
```

@@

```{r geomapNI, eval=FALSE}
CustomNI_map <- geomapNI(Index = CustomNI$CustomIndex, 
                         year = 2019, 
                         OutputType = OutputType, 
                         ecosystem = ecosystem_use,
                         shapeLibraryPath = shapeLibraryPath)
```

@@

```{r}
Index <- CustomNI$CustomIndex
plot(Index$wholeArea, main = "Custom index", cex = 1, lwd = 2, shade = TRUE)
```
```{r}
data.frame("Region"= rownames(summary(Index$wholeArea)), summary(Index$wholeArea), check.names = F) %>%
  flextable() %>%
  theme_vanilla() %>%
  set_table_properties(layout = "autofit")
```
